---
title: "Exam Data Visualisation"
format:
  html:
    toc: true
    number-sections: true
---

```{r, message=FALSE, warning=FALSE} 
library(readr) 
library(ggplot2) 
library(dplyr)
```
This does three things:

- `readr` reads the CSV
- `ggplot2` makes the charts
- `dplyr` helps with simple data handling

------------------------------------------------------------------------

# Step 3: Read the CSV file

Add this next chunk:

```{r} 
exam_data <- read_csv("Exam_data.csv")
```
This assumes your Quarto file and your CSV are in the **same folder**.

If your CSV is in another folder, then you need the correct file path instead.

------------------------------------------------------------------------

# Step 4: Preview the data

Add this:

```{r} 
head(exam_data)
```
This lets you see the first few rows and confirm the file loaded properly.

You can also check the column names:

```{r}
colnames(exam_data)
```
------------------------------------------------------------------------

# Step 5: Make your first simple visualisation

## Bar chart of gender

Now let’s start with the easiest plot.

```{r} 
ggplot(exam_data, aes(x = GENDER)) + geom_bar() + labs( 
  title = "Number of Students by Gender", 
  x = "Gender", 
  y = "Count" )
```
### What this does

-   `ggplot(exam_data, aes(x = GENDER))` says: use `exam_data` and put `GENDER` on the x-axis
-   `geom_bar()` automatically counts how many students are in each gender category
-   `labs()` adds a chart title and axis labels

# What you should expect

This will produce a **bar chart** with: 

- one bar for **Male** 
- one bar for **Female**

Since your dataset has both genders, this is a good first plot to understand category counts.

------------------------------------------------------------------------

# Step 6: Make it look a bit nicer

You can slightly improve the appearance:

```{r}
ggplot(exam_data, aes(x = GENDER, fill = GENDER)) + geom_bar() + labs( 
  title = "Number of Students by Gender", 
  x = "Gender", 
  y = "Count" ) + 
  theme_minimal()
```
### New parts here

-   `fill = GENDER` gives different colours to each gender bar
-   `theme_minimal()` gives a cleaner style

------------------------------------------------------------------------

# Step 7: Render or preview

Now click: 

- **Render** or 
- **Preview** in RStudio.

You should now see your first chart in the output.

------------------------------------------------------------------------

# Step 8: Try a second bar chart

## Bar chart of race

Now use the same idea for `RACE`:

```{r} 
ggplot(exam_data, aes(x = RACE, fill = RACE)) + geom_bar() + labs( 
  title = "Number of Students by Race", 
  x = "Race", 
  y = "Count" ) + 
  theme_minimal()
```
This shows how many students belong to each race category.

------------------------------------------------------------------------

# Step 9: First numerical visualisation

## Histogram of Maths scores

Now let’s visualise a numerical variable.

```{r}
ggplot(exam_data, aes(x = MATHS)) + geom_histogram(binwidth = 5, fill = "steelblue", color = "white") + labs( 
  title = "Distribution of Maths Scores", 
  x = "Maths Score", 
  y = "Frequency" ) + 
  theme_minimal()
```
### What this does

-   `x = MATHS` puts maths marks on the x-axis
-   `geom_histogram()` groups the scores into intervals
-   `binwidth = 10` means the bars cover ranges like 0–10, 10–20, 20–30, and so on

This helps you see: 

-   Where most students’ maths scores lie 
-   Whether the distribution is spread out 
-   Whether there are very low or very high scores

------------------------------------------------------------------------

# Step 10: Another useful plot

## Boxplot of Maths by Gender

Now let’s compare marks across categories.

```{r}
ggplot(exam_data, aes(x = GENDER, y = MATHS, fill = GENDER)) + geom_boxplot(alpha = 0.7) + labs( 
  title = "Maths Scores by Gender", 
  x = "Gender", 
  y = "Maths Score" ) + 
  theme_minimal()
```
### Why this is useful

A boxplot helps you compare: 

-   Median score 
-   Spread of scores 
-   Possible outliers

So instead of just counting students, you are now comparing their performance.

------------------------------------------------------------------------

# Step 11: Scatter plot

## English vs Maths

Now a relationship plot:

```{r} 
ggplot(exam_data, aes(x = ENGLISH, y = MATHS)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "English vs Maths Scores",
    x = "English Score",
    y = "Maths Score"
  ) +
  theme_minimal()
```
This helps answer: 

-   Do students who score higher in English also tend to score higher in Maths? 
-   Is there a visible positive trend?

------------------------------------------------------------------------
