---
title: "Hands-On Exercise 1"
format: html
editor: visual
date: "2026-04-15"
date-modified: "2026-04-15"
---

This exercise looks to explore 

```{r, message=FALSE, warning=FALSE}
library(readr) 
library(ggplot2) 
library(dplyr)
```

This does three things:

-   `readr` reads the CSV
-   `ggplot2` makes the charts
-   `dplyr` helps with simple data handling

------------------------------------------------------------------------

# Read the CSV file

Add this next chunk:

```{r}
exam_data <- read_csv("../data/Exam_data.csv")
```

This assumes your Quarto file and your CSV are in the **same folder**.

If your CSV is in another folder, then you need the correct file path instead.

------------------------------------------------------------------------

# R Graphics VS ggplot

First, we look at R Graphics

```{r}
hist(exam_data$MATHS)
```

Next, we look at ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

------------------------------------------------------------------------

# Geometric Objects

## geom_bar

Now let’s start with the easiest plot.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

Note: y scale is not very useful, in fact it is very misleading.

The following code below does 2 steps:

-   `scale_y_continuous` is used to turn off the y-axis, and
-   binwidth argument is used to change the binwidth to 2.5

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```
## geom_histogram()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

It is to note that the default bin is **30**.

### Modifying a geometric object by changing geom()

The following code chunk does the following:

-   bins argument is used to change the number of bins to 20,
-   fill argument is used to shade the histogram with light blue color, and
-   color argument is used to change the outline colour of the bars in black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

### Modifying a geometric object by changing aes()

The following code changes the fill of the histogram by using sub-group of `aesthetic()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## geom_density()

This computes and plots the kernel density estimate, which is a smoothed version of the histogram.

It is a useflu alternative to the histogram for continuous data that comes from an underlying smooth distribution.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()
```

The next code plots 2 kernel density lines by using colour or fill arguments of `aes()`

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## geom_boxplot()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

**Notches** are used in boxplots to help visually assess whether medians of distributions differ. If it does not overlap, it means that the medians are different.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## geom_violin

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## geom_point()

This is to create scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## Combinatiation of geom objects

The code below shows the plot of data points on the boxplot by using both `geom_boxplot()` and `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

------------------------------------------------------------------------

# Working with stats in ggplot2

## stat_summary()

This adds mean values by using `stat_summary()` and overrides the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## geom()

This adds mean values by using `geom_()` and overrides the default stats.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## Adding best-fit curve on a scatterplot

The code below shows `geom_smooth()` being used to plot a best-fit curve in the scatterplot. The default method used is loess

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

------------------------------------------------------------------------

# Facets

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables.

## facet_wrap()

This code wraps a 1-D sequence of panels into 2-D. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## facet_grid()

This code forms a matrix of panels defined by row and column facetting variables. It is most useful when you have **two discrete variables**, and all combinations of the variables exist in the data.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

------------------------------------------------------------------------

# Coordinates

This group of functions map the position of objects onto the plane of the plot.

## Working with Coordinates

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## coord_flip()

Flips the horizontal bar chart into vertical bar chart

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## Changing the range of x & y-axis

As the scatterplot below is slightly misleading due to the axis range not being equal,

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

the code below fixes both axis range from 0 to 100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

------------------------------------------------------------------------

# Themes

Controlsd the elements of the graph not related to the data, like:

-   background colour
-   font size
-   gridlines
-   colour of labels

The code below plots a horizontal bar chart using `theme_gray()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

This code below plots the bar chart using `theme_classic()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

This code below plots the bar chart using `theme_minimal()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

------------------------------------------------------------------------
