---
title: "Hands-On Exercise 3"
author: "Kelvin Andikko"
date: "2026-04-29"
date-modified: "last-modified"
format: html
editor: visual
---

In this exercise, 2 will be exploring 2 parts of data visualisations.

In the first part of this exercise, we would be exploring the creation of **interactive** data visulisations by using the following packages below:

-   `ggiraph`: makes `ggplot` graphs interactive.
-   `plotly`: R library for plotting interactive statistical graphs.
-   `DT`: provides R interface to the JavaScript library DataTables that create interactive table on the html page
-   `tidyverse`: a family of modern R packages specially designed to support data science, analysis and communication task including the creation of statistical graphs. 
-   `patchwork`: combines multiple ggplot2 graphs into 1 figure

In the second part of this exercise, we would be exploring the creation of **animated** data visualisation by using the following packages below:

-   `plotly`
-   `tidyverse`
-   `gganimate`: an extension of ggplot for creating animated statistical graphs
-   `gifski`: converts video frames to GIF animations using pngquant's fancy features for efficient cross-frame palettes and temporal dithering. It produces animated GIFs that use thousands of colors per frame.
-   `gapminder`: we are using this only for its `country_colors` scheme

The code block belows checks that the above packages are installed and loaded into the R working environment.

```{r, warning=FALSE, message=FALSE}
pacman::p_load(ggiraph, plotly, readxl, gifski, gapminder,
               gganimate, patchwork, DT, tidyverse) 
```

The next code block imports the CSV file:

```{r}
exam_data <- read_csv("../data/Exam_data.csv")
```

------------------------------------------------------------------------

# `ggiraph` - Interactive Data Visualisation

`ggiraph` is an html widget and an extension of `ggplot2`, where it allows the graphics to be interactive.

3 arguments is used to make the graph interactive:

-   **Tooltip**: a column of datasets that contain tooltips to be displayed when the mouse is over elements
-   **Onclick**: a column of data-sets that contain a JavaScript function to be executed when elements are clicked
-   **Data_id**: a column of data-sets that contain an id to be associated with elements

## The `tooltip` aesthetic

The following code chunks below plots the interactive statistical graph using the `ggiraph` pacakge. This performs the following:

-   `ggplot()` object is first created. along with `geom_dotplot_interactive()` where it generates an interactive version of the graph
-   `girafe()` is used to generate an svg object to be displayed on an html page

### Tooltip with only Student's ID

```{r,warning=FALSE, message=FALSE}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```
When the mouse hovers over the points in the graph above, it will display the student's ID. This is done by editing in the `geom_dotplot_interactive()` function, where `aes(tooltip = ID)`.

### Displaying multiple information using tooltip

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```
In this code block, where both the student's ID and the class the student is in are to be displayed, a variable `exam_data$tooltip` is created in the first 3 lines of code, where it defines how it will be displayed. Afterwards, in the main body of code, tooltip is then assigned to `exam_data$tooltip`.

### Customise your tooltip style

```{r}
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = ID),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)                                        
```

In this code block, the style is changed, where the background colour is while and the font colour and black and bolded.

This is done by introducing the variable `tooltip_css`, where you will assign the background colour, font style and the colour.

Afterwards, within the `girafe()` function, there is an additional `options`, where you would reference the `tooltip_css` as such there.

### Advanced tooltip: displaying statistics

Let's say that I want to display the **90% confidence interval of the mean** of each subjects in the graph below.

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

In the code chunk above, `tooltip` is referenced in the first 4 lines, where the function is created to calculate the mean scores for the subjects with the intervals, along with how it is going to be displayed.

After that, it is then pointed within `stat_summary()` as shown in the code chunk.

------------------------------------------------------------------------

## The `data_id` aesthetic

Let's say, I want to identify visually the students in the same class over the histodot graph below.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(           
    aes(data_id = CLASS),             
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)                                        
```

What is done is that `data_id` is used within `geom_dotplot_interactive()` where it referenced CLASS (`data_id = CLASS`). As such, when the clicker hovers over the dots, the dots will change colour to orange, where it indicates the students that are in the same class. This allows me to visually see the distributions of the students.

It is to note that the default colour for this is orange, where:
`hover_css = "fill:orange;"`

### Re-styling the hover effect

Instead of the dot's colour changing to orange, I would want to highlight the students that are in the same class.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

In the code block above, `options` is introduced within the `girafe()` function, where you can do the highlighting effect as such.

------------------------------------------------------------------------

## Combine both tooltip and the hover effect

Now, I want to combine both effects on the statistical graph. The code chunk below shows how it is done.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

The graph shows how the students in the same class is both highlighted, and shown as the tooltip.

------------------------------------------------------------------------

## The `onclick` effect

This argument provides a way where upon the click of an data point, it leads you to another webpage.

This code chunk shows an example of the use of the argument, where `exam_data$onclick` is introduced in the first 3 lines.

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)                                        
```


------------------------------------------------------------------------

## Coordinated Multiple Views

In hands-on exercise 2, we came across the use of multiple graphs. This can also be applied in interactive statistical graphs as well.

In this example, I will be displaying the graph of marks for both Maths and Science. When the mouse hovers over a data point in either graph, the corresponding data point on the other visualisation will be highlighted as well.

This is done by:

-   Using appropriate interactive functions of `ggiraph` to create the multiple views.
-   Using `patchwork` function within the `girafe` function to create the interactive coordinated multiple views.

The code chunk below showcases that.

```{r}
p1 <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```

The `data_id` aesthetic is critical to link observations between both plots, while the tooltip aesthetic is *optional* but a nice to have to improve the visualisation.

------------------------------------------------------------------------

# `plotly` - an alternative to Interactive Data Visualisation

This library creates interactive web graphics from `ggplot2` graphs and/or a custom interface to the (MIT-licensed) JavaScript library `plotly.js` inspired by the grammar of graphics.

There are 2 ways of using plotly to create interactive graph:

-   by using `plot_ly()`, or
-   by using `ggplotly()`

## The `plot_ly()` method

The graph below shows an example of a basic interactive plot created using `plot_ly()`.

```{r, warning=FALSE, message=FALSE}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```

### Adding visuals to `plot_ly()`

The *colour* argument is mapped to a qualitative visual variable (i.e. RACE)

```{r, warning=FALSE, message=FALSE}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

## The `ggplotly()` method

The graph below shows an example of a basic interactive plot created using `ggplotly()`.

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```

The only extra line to add is `ggplotly()`!

## Coordinated Multiple Views with plotly

To create a coordinated linked plot using plotly, 3 steps need to be done:

-   `highlight_key()` of plotly is used as shared data
-   2 scatterplots will be created using `ggplot2` functions
-   `subplot() of plotly is used to place them next to each other side-by-side

```{r}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```

It is to note that `highlight_key()` simply creates an object of class `crosstalk::SharedData`.

------------------------------------------------------------------------

# The `crosstalk` Way of Interactive Data Visualisation

This is an add-on to the htmlwidgets packages, where it extends it with a set of classes, functions, and conventions over implementing cross-widget interactions.

## The Data Table, aka DT

This is a wrapper of the JavaScript Library **DataTables**.

Data objects in R can be rendered as HTML tables using this library.

```{r}
DT::datatable(exam_data, class= "compact")
```

## The `crosstalk` method - Linked Brushing

The following code chunk below implements the coordinated brushing.

```{r}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```

The `highlight()` is a function of the plotly package, where it sets a variety of options for brushing (basically highlighting) multiple plots. It is primarily designed for linking multiple plotly graphs, and it may not behave as expected when linking plotly to another htmlwidget package via crosstalk.

The `bscols()` is a helper function of crosstalk package, where it makes it easy to input HTML elements side by side. It can be called directly from the console but it is especially designed to work in a R markdown document. **WARNING:** This will bring in all of Bootstrap!

------------------------------------------------------------------------

# Making statistical graphs animated

While telling a visually-driven story about your data, often animation value adds by attracting interests of the audience, and brings the wow factor, making deeper impressions.

In this part of the exercise. we will use both `gganimate` and `plotly` packages, along with `tidyr`, which is used to reshape data and `dplyr`, which is used to process, wrangle and transform data.

## Importing the dataset

First, the code chunk below imports the data *GlobalPopulation* where it will be used for this exercise.

```{r, warning=FALSE, message=FALSE}
col <- c("Country", "Continent")
globalPop <- read_xls("../data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate(across(col, as.factor)) %>%
  mutate(Year = as.integer(Year))
```

------------------------------------------------------------------------

## `gganimate` - The ANImethod Data Visualisation

This extends the grammar of graphics as implemented by ggplot2 to include the description of animation. It does this by providing a range of new grammar classes that can be added to the plot object in order to customise how it should change with time.

The functions are as shown below:

-   `transition_*()`: how the data should be spread out and how it relates to itself across time
-   `view_*()`: how the positional scales should change along the animation
-   `shadow_*`()`: how data from other points in time should be presented in the given point in time
-   `enter_*()/exit_*()`: how new data should appear and how old data should disappear during the course of the animation
-   `ease_aes()`: how different aesthetics should be eased during transitions


### Creating a static population bubble plot

The code chunk below creates a static bubble plot:

```{r, warning=FALSE, message=FALSE}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') 
```


### Building it up to be animated

The following code chunk does the following:

-   `transition_time()` of **gganimate** will be used to create the transition through distinct states in time, which in this case, its year.
-   `ease_aes()` is used to control easing of aesthetics. By default, it is linear. Other methods include quadratic, cubic, quartic, quintic, sine, circular, exponential, elastic, back, and bounce.

```{r, warning=FALSE, message=FALSE}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')          
```

------------------------------------------------------------------------

## `plotly` - The Other ANImethod of Data Visualisation

In this package, both `ggplotly()` and `plot_ly()` support key frame animations through the frame argument/aethetics. They also support an ids argument/aesthetic to ensure smooth transitions between objects with the same id (which helps facilitate object constancy).

### Building the animation with `ggplotly()`

In the code block below, the following is done:

-   Appropriate ggplot2 functions are used to create a static bubble plot. The output is then saved as an R object called gg.
-   `ggplotly()` is then used to convert the R graphic object into an animated svg object

In this example, if I want to not display the legend, the following  would be used: `show.legend = FALSE`. However, `theme(legend.position='none')` is used, as the legend would still appear if the previous argument was used.

```{r, warning=FALSE, message=FALSE}
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position='none')

ggplotly(gg)
```

### Building the animation with `plot_ly()`

The code chunk below shows how `plot_ly()` is utilised to animate the bubble plot.

```{r, warning=FALSE, message=FALSE}
bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```

------------------------------------------------------------------------